`traveltime`: an R package to calculate walking and motorised travel time from user-specified locations

### 2025-01-24 paper/paper.qmd
fix citation format so referenes work resolve #42

---
title: "`traveltime`: an R package to calculate walking and motorised travel time from user locations"
format:
  #md: default
  html:
    keep-md: true
    fig-height: 4
    fig-align: center
    fig-format: png
    dpi: 300
tags: 
  - R
  - geographic information systems
  - spatial analysis
authors:
  - name: Gerard E Ryan
    orcid: 0000-0003-0183-7630
    corresponding: true
    affiliation: "1, 2"
  - name: Nick Tierney
    orcid: 0000-0003-1460-8722
    affiliation: "1, 3"
  - name: Nick Golding
    orcid: 0000-0001-8916-5570
    affiliation: "1, 3 "
  - name: Daniel Weiss
    orcid: 0000-0002-6175-5648
    affiliation: "1, 3"
affiliations:
 - name: The Kids Research Institute Australia, Nedlands 6009 WA, Australia
   index: 1
 - name: Melbourne School of Population and Global Health, University of Melbourne, 3010, VIC, Australia
   index: 2
 - name: Curtin University, Bentley, WA, Australia
   index: 3
date: In prep 2025
bibliography: paper.bib
---

@@

Understanding and mapping the time to travel among locations is useful for many activities from urban planning `[@zahavi1974traveltime]` to public health `[@weiss2020global]` and likely a myriad others we haven't thought of. Here we present a software package --- `traveltime` --- written in and for the language R `[@Rref]`. `traveltime` enables a user to supply a set of geographic point locations and an extent of interest, and calculate the motorised or walking travel time over that extent. The result is a raster of 30 arcsecond resolution (approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator) where the value in each cell is the lowest travel time in minutes to any of the supplied point locations over the extent of interest.

@@

`traveltime` provides a spatial interface using object classes from the `terra` package `[@terra]`. It accepts points as either `data.frame` or `SpatVector` class objects; extents as `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` objects; and return the results as a `SpatRaster` object. The package uses `gdistance` `[@gdistance2017]` to calculate the minimum least-cost-distance for each cell from the points of interest.The travel time is calculated as movement over a 'friction surface' based-on the surfaces generated by `@weiss2020global` available via the `malariaAtlas` package `[@pfeffer2018malariaatlas]`. Rather than house and manipulate these surfaces independently, `traveltime` relies on `malariaAtlas` to download the surfaces for the extent of interest.

@@

Reference documents for `traveltime` are available at <https://idem-lab.github.io/traveltime/>. Although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package should be accompanied by citing the underlying work by `@weiss2018global; @weiss2020global`.

@@

Global maps of travel time to cities `[@weiss2018global]` and health care facilities `[@weiss2020global]` `@fig-tt_healthcare` have generated significant interest and use. Collectively `@weiss2018global ; @weiss2020global` have over 1300 citations (Google Scholar, at 10th December 2024), and the city data set is available to R users through the widely-used `geodata` package `[@geodata]`. There is clear demand for these type of products.

`@weiss2020global` and colleagues made their code available as an R script to allow for reproduction and extension of their analyses (<https://malariaatlas.org/wp-content/uploads/2022/11/R_generic_accessibilty_mapping_script_2020-1.txt> ). To further enable extension of this work, here we have developed an R package based on that code to seamlessly calculate the travel time from any arbitrary set of locations.

Other R packages produce superficially similar though fundamentally different functionality. A gaggle of R packages provide interfaces to the <https://www.TravelTime.com> API `[@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh]`. The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons, which are areas reachable within a given time from a given location. The isochron polygons are most comparable to what the `traveltime` package calculates, though each calculation is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than gradation across cells. TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repetitive iteration for all combinations of time and points, and additional calculations. Furthermore, TravelTime.com requires access keys, paid subscription beyond a short free period, caps queries, which add considerable friction to the use experience.

@@

We can download a national-level polygon of Singapore from the GADM `[@gadm]` database using the `geodata` package `[@geodata]`. Here we download only the national boundary (`level = 0`) and at a low resolution (`resolution = 2`).

@@

The the `stations` data set included in the `traveltime` package is a 563 \$\*\$ 2 matrix containing the longitude (x) and latitude (y) of all LRT and MRT station exits in Singapore from `@singdata:2024`.

@@

We can pass in our basemap `sin` `@fig-basemap`, a `SpatVector`, directly as the `extent`. It would also be possible pass in a `SpatRaster` with extent equal to our area of interest, or specify coordinates either as a vector, matrix, or `SpatExtent`.

@@

>>>>

---
title: "`traveltime`: an R package to calculate walking and motorised travel time from user-specified locations"
format:
  #md: default
  html:
    keep-md: true
    fig-height: 4
    fig-align: center
    fig-format: png
    dpi: 300
tags: 
  - R
  - geographic information systems
  - spatial analysis
authors:
  - name: Gerard E Ryan
    orcid: 0000-0003-0183-7630
    corresponding: true
    affiliation: "1, 2"
  - name: Nick Tierney
    orcid: 0000-0003-1460-8722
    affiliation: "1, 3"
  - name: Nick Golding
    orcid: 0000-0001-8916-5570
    affiliation: "1, 3 "
  - name: Daniel Weiss
    orcid: 0000-0002-6175-5648
    affiliation: "1, 3"
affiliations:
 - name: The Kids Research Institute Australia, Nedlands 6009 WA, Australia
   index: 1
 - name: Melbourne School of Population and Global Health, University of Melbourne, 3010, VIC, Australia
   index: 2
 - name: Curtin University, Bentley, WA, Australia
   index: 3
date: In prep 2025
bibliography: paper.bib
---

@@

Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@weiss2020global] and likely a myriad others we haven't thought of. Here we present a software package --- `traveltime` --- written in and for the language R [@Rref]. `traveltime` enables a user to supply a set of geographic point locations and an extent of interest, and calculate the motorised or walking travel time over that extent. The result is a raster of 30 arcsecond resolution (approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator) where the value in each cell is the lowest travel time in minutes to any of the supplied point locations over the extent of interest.

@@

`traveltime` provides a spatial interface using object classes from the `terra` package `[@terra]`. It accepts points as either `data.frame` or `SpatVector` class objects; extents as `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` objects; and return the results as a `SpatRaster` object. The package uses `gdistance` [@gdistance2017] to calculate the minimum least-cost-distance for each cell from the points of interest. The travel time is calculated as movement over a 'friction surface' based-on the surfaces generated by @weiss2020global available via the `malariaAtlas` package [@pfeffer2018malariaatlas]. Rather than house and manipulate these surfaces independently, `traveltime` relies on `malariaAtlas` to download the surfaces for the extent of interest.

@@

Reference documents for `traveltime` are available at <https://idem-lab.github.io/traveltime/>. Although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package should be accompanied by citing the underlying work by @weiss2018global; @weiss2020global.

@@

Global maps of travel time to cities [@weiss2018global] and health care facilities [@weiss2020global] `@fig-tt_healthcare` have generated significant interest and use. Collectively @weiss2018global ; @weiss2020global have over 1300 citations (Google Scholar, at 10th December 2024), and the city data set is available to R users through the widely-used `geodata` package [@geodata]. There is clear demand for these type of products.

@weiss2020global and colleagues made their code available as an R script to allow for reproduction and extension of their analyses (<https://malariaatlas.org/wp-content/uploads/2022/11/R_generic_accessibilty_mapping_script_2020-1.txt> ). To further enable extension of this work, here we have developed an R package based on that code to seamlessly calculate the travel time from any arbitrary set of locations.

Other R packages produce superficially similar though fundamentally different functionality. A gaggle of R packages provide interfaces to the <https://www.TravelTime.com> API [@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh]. The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons, which are areas reachable within a given time from a given location. The isochron polygons are most comparable to what the `traveltime` package calculates, though each calculation is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than gradation across cells. TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repetitive iteration for all combinations of time and points, and additional calculations. Furthermore, TravelTime.com requires access keys, paid subscription beyond a short free period, caps queries, which add considerable friction to the use experience.

@@

We can download a national-level polygon of Singapore from the GADM [@gadm] database using the `geodata` package [@geodata]. Here we download only the national boundary (`level = 0`) and at a low resolution (`resolution = 2`).

@@

The the `stations` data set included in the `traveltime` package is a 563 \$\*\$ 2 matrix containing the longitude (x) and latitude (y) of all LRT and MRT station exits in Singapore from @singdata:2024.

@@

We can pass in our basemap `sin` @fig-basemap, a `SpatVector`, directly as the `extent`. It would also be possible pass in a `SpatRaster` with extent equal to our area of interest, or specify coordinates either as a vector, matrix, or `SpatExtent`.

@@

GR is supported by the Gates Foundation grant for the Vector Atlas.
